---
title: "A549_spheroids_analysis"
format:
  html:
    toc: true
    number-sections: true
execute:
  cache: false
  freeze: auto
editor: 
  markdown: 
    wrap: 72
---

## Package & object/signatures loading

```{r}
#Load Package
library(Matrix)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(cowplot)
library(dyno)
library(dynwrap)
library(tislingshot)
library(SingleCellExperiment)
library(hdf5r)
library(hexbin)
library(tibble)
library(dplyr)
library(fgsea)
library(msigdbr)
library(reshape2)
library(ComplexHeatmap)
library (tidyr)
library(colorRamp2)
library(readxl)
library(VennDiagram)
library(grid)
library(clusterProfiler)
library(enrichplot)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(ggVennDiagram)
library(babelwhale)
library(openxlsx2)
library(AUCell)
library(svglite)

# INPUT : Load the A549 spheroid seurat object 
data <- readRDS(file = "/Users/fauveauc/Documents/R_studio/A549 analysis/seurat_object_A549_pers_final.rds")

# INPUT : Obtain signatures from the supplementary tables of the publication :
  #Load the excel file
df <- read_excel("/Users/fauveauc/Documents/R_studio/A549 analysis/Supplemental_tables.xlsx", sheet = "Table S.7")
  # transfrom it into a list containing the different signatures 
Signatures <- lapply(seq_len(nrow(df)), function(i) {
  g <- unlist(df[i, -1], use.names = FALSE)
  g <- as.character(g)
  g <- trimws(g)
  g <- g[!is.na(g) & g != ""]
  unique(g)
})
names(Signatures) <- df[[1]]

# OUTPUT: Determine the path to the directory where the figures and table will be saved
path_to_save <- "/Users/fauveauc/Documents/R_studio/A549 analysis/figures_and_tables"


rm(df)
```

## Dataset overview

### UMAPs

```{r}
DimPlot(data, reduction = "umap.harmony", group.by = "treatment")
ggsave("Fig_2A.png", path = path_to_save)
DimPlot(data, reduction = "umap.harmony", group.by = "harmony_clusters", label = T)
ggsave("Fig_2B.png", path = path_to_save)
```

### Distribution per cluster

```{r}

# Samples distribution within the different clusters
counts <- group_by(data@meta.data, sample, harmony_clusters) %>% summarise(count = n())
Percentage_Cluster<- counts %>% group_by(harmony_clusters) %>% mutate(ValuePer=(count/sum(count)*100)) %>% ungroup()

ggplot(Percentage_Cluster, aes(harmony_clusters,ValuePer,fill=sample))+
 geom_bar(stat = 'identity')


# Treatment distribution within the different clusters
counts <- group_by(data@meta.data, treatment, harmony_clusters) %>% summarise(count = n())
Percentage_Cluster<- counts %>% group_by(harmony_clusters) %>% mutate(ValuePer=(count/sum(count)*100)) %>% ungroup()

ggplot(Percentage_Cluster, aes(harmony_clusters,ValuePer,fill=treatment))+
 geom_bar(stat = 'identity')


# Biological replicates distribution within the different clusters
counts <- group_by(data@meta.data, replicate, harmony_clusters) %>% summarise(count = n())
Percentage_Cluster<- counts %>% group_by(harmony_clusters) %>% mutate(ValuePer=(count/sum(count)*100)) %>% ungroup()

ggplot(Percentage_Cluster, aes(harmony_clusters,ValuePer,fill=replicate))+
 geom_bar(stat = 'identity')
ggsave("Fig_S1C.png", path = path_to_save)


rm(Percentage_Cluster, counts)
```

### Clusters markers

```{r}
Idents(data)<-"harmony_clusters"
DEG_clusters_data<- FindAllMarkers(data, only.pos = TRUE, logfc.threshold = 0.25, min.pct = 0.3)
DEG_clusters_data<- dplyr::filter(DEG_clusters_data, p_val_adj<0.05)

#Save the list of markers for each clusters 
write_xlsx(DEG_clusters_data, file.path(path_to_save, "Table_S1.xlsx"))

top3_markers <- data.frame()

for (cluster in levels(data))
{
b <- arrange(DEG_clusters_data[DEG_clusters_data$cluster == cluster,],p_val_adj -avg_log2FC,-pct.1)
if(nrow(b) > 3) {b<- b[1:3,] }
top3_markers <- rbind(top3_markers, b)
}
rm(b)

DotPlot(data, assay = 'RNA',features = top3_markers$gene, dot.scale = 5) + RotatedAxis()
ggsave("Fig_S1D.png", path = path_to_save, width=11 )

rm(DEG_clusters_data,top3_markers,cluster)
```

## Cell cycle Analysis

```{r}
# cell cycle scoring
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

data <- CellCycleScoring(data, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

#Visualization
DimPlot(data, group.by = "Phase", reduction = "umap.harmony", label = FALSE, pt.size = 0.7)
ggplot(data@meta.data, aes(x=treatment, fill=Phase)) + geom_bar(position = "fill")

rm(g2m.genes, s.genes)
```

## Trajectory

### Trajectory generation

```{r}

#The dataset is created with the expression matrix, the grouping variable (treatment) and the dimensionality reduction coordinates (UMAP) from the Seurat object. The expression matrix is transposed to have cells as rows and genes as columns, which is the format expected by dyno.
dataset <- wrap_expression(counts = t(data@assays$RNA$counts), expression = t(data@assays$RNA$data))
dataset <- add_grouping(dataset, as.character(data$treatment))
dataset <- add_dimred(dataset, data@reductions$umap.harmony@cell.embeddings)
dataset$grouping

#The trajectory is inferred using the Slingshot method
model <- infer_trajectory(dataset, method = tislingshot::ti_slingshot())



# Milestones are renamed with letters
get_milestone_labelling(model)
new_names <- c("2" = "A", "1" = "B", "3" = "C", "4" = "D")
model <- label_milestones(model, new_names)
get_milestone_labelling(model)


#Visualization
plot_dimred(model, color_cells = "auto", size_transitions = 1, label_milestones = TRUE)

plot_dimred(model, dimred = dataset$dimred, color_cells = "pseudotime", 
            grouping = dataset$grouping, size_transitions = 1, label_milestones = TRUE)

plot_dimred(model, dimred = dataset$dimred, color_cells = "milestone", 
            grouping = dataset$grouping, size_transitions = 1, label_milestones = TRUE)
ggsave("Fig_2D.png", path = path_to_save,height = 5)

rm(new_names)
```

### Heatmap of important features

```{r}
#Calculation of the an importance score for each gene (reflects gene contribution to the trajectory)
feature_imp <- calculate_overall_feature_importance(model, expression_source = dataset$expression)

#Save the table of genes with their importance score
write_xlsx(feature_imp, file.path(path_to_save, "Table_S2.xlsx"))

#Heatmap of the 20 most important genes
top_features <- feature_imp %>% 
  top_n(20, importance) %>% 
  pull(feature_id)
plot_heatmap(
  model, 
  expression_source = dataset$expression, 
  features_oi = top_features
)
ggsave("Fig_2E.png", path = path_to_save, height = 4,width = 4)

rm(top_features, feature_imp)
```

## UMAP of NRF2 and Hypoxia

```{r}
# Extract the expression matrix 
expr_mat <- GetAssayData(data, assay = "RNA")

# Build AUCell rankings
expr_ranks <- AUCell_buildRankings(expr_mat)

# Calculate AUCell score for the pathway gene set
nrf2_auc <- AUCell_calcAUC(list(NRF2 = Signatures[["NRF2_SIGNATURE_NAMANI"]]), expr_ranks)
hypoxia_auc <- AUCell_calcAUC(list(hypoxia = Signatures[["HALLMARK_HYPOXIA"]]), expr_ranks)

# Extract the score for each cell
nrf2_scores <- as.numeric(getAUC(nrf2_auc)["NRF2", ])
hypoxia_scores <- as.numeric(getAUC(hypoxia_auc)["hypoxia", ])

# Add the score to Seurat metadata
data$NRF2score <- nrf2_scores
data$hypoxiascore <- hypoxia_scores

# Plot on UMAP
FeaturePlot(
  data,
  features = "hypoxiascore",
  reduction = "umap.harmony"
)+ scale_color_gradientn(colors = c("grey90", "#B3D9FF", "#0073E6", "#001F4D"))
ggsave("Fig_2F1.png", path = path_to_save, width=4, height = 3)
FeaturePlot(
  data,
  features = "NRF2score",
  reduction = "umap.harmony"
)+ scale_color_gradientn(colors = c("lightgrey", "pink", "red", "darkred"))
ggsave("Fig_2F2.png", path = path_to_save, width=4, height = 3)

rm(expr_mat, expr_ranks,nrf2_auc, nrf2_scores, hypoxia_auc, hypoxia_scores)
```

### Violin plot of NRF2-associated genes across milestones

#### Creation of a meta-data column attributing a milestone state to each cells

```{r}
# Get milestone score from each cells
milestone_info <- model$milestone_percentages

# Identify the milestone with the highest association for each cell
cell_milestone_mapping <- milestone_info %>%
  group_by(cell_id) %>%
  top_n(1, wt = percentage) %>%
  ungroup()

milestone_vector <- setNames(cell_milestone_mapping$milestone_id, cell_milestone_mapping$cell_id)

# Check if all cell IDs exist in the Seurat object
common_cells <- intersect(rownames(data@meta.data), names(milestone_vector))

# Add milestone information as metadata in the Seurat object
data <- AddMetaData(data, metadata = milestone_vector[common_cells], col.name = "Milestone")

data$Milestone <- recode(data$Milestone,   `1` = "B",  `2` = "A" , `3` = "C",  `4` = "D")

# Check the updated milestones
table(data$Milestone)

rm(milestone_info,milestone_vector,common_cells, cell_milestone_mapping)
```

#### Violin plot

```{r}

#Plot of NRF2-associated genes across the different milestones
VlnPlot(data, features = c("NQO1","FTL","CPLX2","EPHX1", "ALDH3A1"), group.by = "Milestone", stack = TRUE, flip = TRUE, split.by = "Milestone", cols= c("#FA8A7D","#F6FBB4", "#A0CBCF","#C2BEDC"))
ggsave("Fig_S2C.png", path = path_to_save)

#Plot of the NRF2 score generated with AUCell across the different milestones
VlnPlot(data, features = "NRF2score", group.by = "Milestone", flip = TRUE, cols= c("#FA8A7D","#F6FBB4", "#A0CBCF","#C2BEDC"),pt.size = 0)
ggsave("Fig_2G.png", path = path_to_save, width = 7)

#Statistical comparison of the NRF2 score across the different milestones
kruskal.test(NRF2score ~ Milestone, data = data@meta.data)
pairwise.wilcox.test(data$NRF2score,data$Milestone,p.adjust.method = "BH")

rm(model,dataset)
```


## Pathway analysis

### Heatmap of Hallmark pathways in clusters + treatment/cell cycle/milestone distribution

#### Run the GSEA

```{r}
# Generate the DEG for each clusters
Idents(data)<-"harmony_clusters"
markers_clusters<- FindAllMarkers(data,  min.pct = 0.3)


# Generate the gene list for each cluster, ranked by log2FC
ranked_gene_lists <- markers_clusters %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC)) %>%
  group_split() %>%
  setNames(unique(markers_clusters$cluster))

# Convert it to named vectors for fgsea
ranked_lists <- lapply(ranked_gene_lists, function(df) {
  setNames(df$avg_log2FC, df$gene)  # Named vector with logFC values
})

# Load MSigDB gene sets
m_df <- msigdbr(species = "Homo sapiens", category = "H")
fgsea_sets <- m_df %>% split(x = .$gene_symbol, f = .$gs_name)


# Run GSEA for each cluster
gsea_results <- lapply(names(ranked_lists), function(cluster) {
  fgsea(pathways = fgsea_sets, stats = ranked_lists[[cluster]])
})
names(gsea_results) <- names(ranked_lists)

# Convert results to a data frame
gsea_results_df <- do.call(rbind, Map(cbind, gsea_results, cluster = names(gsea_results)))



rm(m_df, markers_clusters, ranked_gene_lists, ranked_lists, gsea_results)
```

#### GSEA results preparation for the heatmap

```{r}

#create a data frame with patwhays and pvalues & removes pathway not significantly modified in at least one cluster and values with NA
heatmap_data_pvalue <- dcast(gsea_results_df, pathway ~ cluster, value.var = "padj", fill = 1)
heatmap_data_pvalue<- heatmap_data_pvalue[apply(heatmap_data_pvalue[, -1], 1, function(x) any(x < 0.01)), ]


#create a data frame with patwhays and NES 
heatmap_data <- dcast(gsea_results_df, pathway ~ cluster, value.var = "NES", fill = 0)
#removes pathway not significantly enriched in at least one clusters (using the heatmap_data_pvalue)
heatmap_data <- heatmap_data[heatmap_data$pathway %in% heatmap_data_pvalue$pathway, ]

# Filter out pathways whose NES is not superior to 1 in at least one cluster
heatmap_data <- heatmap_data[apply(heatmap_data[, -1], 1, function(x) any(x > 1)), ]
heatmap_data_pvalue <- heatmap_data_pvalue[heatmap_data_pvalue$pathway %in% heatmap_data$pathway, ]

# Convert to matrix
heatmap_matrix_NES <- as.matrix(heatmap_data[, -1])
rownames(heatmap_matrix_NES) <- heatmap_data$pathway
heatmap_matrix_pval <- as.matrix(heatmap_data_pvalue[, -1])
rownames(heatmap_matrix_pval) <- heatmap_data_pvalue$pathway

rm(heatmap_data, heatmap_data_pvalue, gsea_results_df)
```

#### Preparation of cell cycle distribution per clusters

```{r}

# Create a table of counts for each cluster-phase combination
counts_df <- data@meta.data %>%
  group_by(harmony_clusters, Phase)%>%
  tally()%>%
  ungroup() %>%
  mutate(harmony_clusters = as.numeric(as.character(harmony_clusters)))%>%
  arrange(harmony_clusters)

# Reshape the data to have clusters as rows and phases as columns
cell_repartition_matrix <- counts_df %>%
  spread(key = Phase, value = n, fill = 0)

# Calculate the proportions by dividing by the row sum (cluster total)
cell_repartition_matrix_prop <- cell_repartition_matrix %>%
  mutate(total_cells = rowSums(across(-harmony_clusters)),  
         across(-c(harmony_clusters, total_cells), ~ . / total_cells, .names = "prop_{.col}"))  

 cell_repartition_matrix_prop <- cell_repartition_matrix_prop[, !names(cell_repartition_matrix_prop) %in% c("G1", "S", "G2M","total_cells")]

# Ensure the matrix is numeric while keeping row and column names
cell_repartition_matrix_prop <- as.matrix(cell_repartition_matrix_prop)  
storage.mode(cell_repartition_matrix_prop) <- "numeric"  

# Transpose the matrix
cell_repartition_matrix_transposed <- t(cell_repartition_matrix_prop)
colnames(cell_repartition_matrix_transposed) <- as.character(cell_repartition_matrix_transposed[1, ])
cell_repartition_matrix_transposed <- cell_repartition_matrix_transposed[-1, , drop = FALSE]
cell_repartition_matrix_prop <- t(cell_repartition_matrix_transposed)

# Prepare data for the heatmap
hac = HeatmapAnnotation(cell_cycle = anno_barplot(cell_repartition_matrix_prop, gp = gpar(fill= c("#F8766D","#00BA38","#619CFF")),  bar_width = 1))


```

#### Preparation of control/treated cell distribution per clusters

```{r}

# Create a table of counts for each cluster-treatment combination
counts_df <- data@meta.data %>%
  group_by(harmony_clusters, treatment) %>%
  tally() %>%
  ungroup() %>%
  mutate(harmony_clusters = as.numeric(as.character(harmony_clusters))) %>%
  arrange(harmony_clusters)


# Reshape the data to have clusters as rows and treatment as columns
cell_repartition_matrix <- counts_df %>%
  spread(key = treatment, value = n, fill = 0)

# Calculate the proportions by dividing by the row sum (cluster total)
cell_repartition_matrix_prop <- cell_repartition_matrix %>%
  mutate(total_cells = rowSums(across(-harmony_clusters)),  
         across(-c(harmony_clusters, total_cells), ~ . / total_cells, .names = "prop_{.col}"))  
 cell_repartition_matrix_prop <- cell_repartition_matrix_prop[, !names(cell_repartition_matrix_prop) %in% c("control", "treated","total_cells")]

# Transpose the matrix
cell_repartition_matrix_transposed <- t(cell_repartition_matrix_prop)
colnames(cell_repartition_matrix_transposed) <- as.character(cell_repartition_matrix_transposed[1, ])
cell_repartition_matrix_transposed <- cell_repartition_matrix_transposed[-1, , drop = FALSE]
cell_repartition_matrix_prop <- t(cell_repartition_matrix_transposed)

# Prepare data for the heatmap
hat = HeatmapAnnotation(treatment = anno_barplot(cell_repartition_matrix_prop, gp = gpar(fill= c("#F8766D","#00BFC4")),  bar_width = 1))

```

#### Preparation of milestone distribution per clusters

```{r}

# Create a table of counts for each cluster-milestone combination
counts_df <- data@meta.data %>%
  group_by(harmony_clusters, Milestone) %>%
  tally() %>%
  ungroup() %>%
  mutate(harmony_clusters = as.numeric(as.character(harmony_clusters))) %>%
  arrange(harmony_clusters)


# Reshape the data to have clusters as rows and treatment as columns
cell_repartition_matrix <- counts_df %>%
  spread(key = Milestone, value = n, fill = 0)

# Calculate the proportions by dividing by the row sum (cluster total)
cell_repartition_matrix_prop <- cell_repartition_matrix %>%
  mutate(total_cells = rowSums(across(-harmony_clusters)),  
         across(-c(harmony_clusters, total_cells), ~ . / total_cells, .names = "prop_{.col}"))  
 cell_repartition_matrix_prop <- cell_repartition_matrix_prop[, !names(cell_repartition_matrix_prop) %in% c("A", "B","C","D","total_cells")]

# Transpose the matrix
cell_repartition_matrix_transposed <- t(cell_repartition_matrix_prop)
colnames(cell_repartition_matrix_transposed) <- as.character(cell_repartition_matrix_transposed[1, ])
cell_repartition_matrix_transposed <- cell_repartition_matrix_transposed[-1, , drop = FALSE]
cell_repartition_matrix_prop <- t(cell_repartition_matrix_transposed)

# Prepare data for the heatmap
ham = HeatmapAnnotation(Milestone = anno_barplot(cell_repartition_matrix_prop, gp = gpar(fill= c("#FA8A7D","#F6FBB4", "#A0CBCF","#C2BEDC")),  bar_width = 1))

rm(cell_repartition_matrix, counts_df,cell_repartition_matrix_prop,cell_repartition_matrix_transposed)

```

#### Heatmap generation

```{r}

svglite(file.path(path_to_save, "Fig_2C.svg"), width = 7, height = 7)

Heatmap(
  heatmap_matrix_NES,
  name = "NES",
  col = colorRamp2(c(min(heatmap_matrix_NES), 0, max(heatmap_matrix_NES)), c("blue", "white", "red")),
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_column_names = TRUE,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 8),  # Adjust row name size
  row_names_max_width = unit(20, "cm"),
  heatmap_legend_param = list(
    title = "NES",
    labels_gp = gpar(fontsize = 10, fontface = "italic"),  # Adjust legend label font
    legend_width = unit(15, "cm")   # Adjust legend width
  ),
    cell_fun = function(j, i, x, y, w, h, fill) {
   # Check for NA values first
    
      # Add asterisks based on p-value thresholds
      if (heatmap_matrix_pval[i, j] < 0.001) {
        grid.text("∗∗∗", x, y, gp = gpar(fontsize = 13, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.01) {
        grid.text("∗∗", x, y, gp = gpar(fontsize = 13, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.05) {
        grid.text("∗", x, y, gp = gpar(fontsize = 13, col = "black"))
      
    }
  },
  top_annotation = c(hat, hac,ham))# Combine the annotations

dev.off()

rm(hac, ham, hat, heatmap_matrix_NES, heatmap_matrix_pval)

```

### Enrichment plots
#### Generation of the ranked gene list for GSEA

```{r}
#Generate the DEG list 
Idents(data)<-"treatment"
markers<- FindMarkers(data,ident.1 = "treated" ,ident.2 = "control" ,logfc.threshold = 0, min.pct = 0.3)
markers$gene<- rownames(markers) 

#Save the table of deregulated genes 
write_xlsx(markers, file.path(path_to_save, "Table_S3.xlsx"))

#Verification of the gene order & comparison sens
markers$stat <- markers$avg_log2FC 
markers %>% arrange(desc(stat), desc(stat)) %>% head(10)
Idents(data) <- data@meta.data$treatment
FeaturePlot(data, features = c("ABCA12"), keep.scale = "feature", pt.size = 1.2,
            label = TRUE, reduction="umap.harmony", order = T)


#Rank  table and create a list with the gene and their rank
markers$gene <- rownames(markers)
ranks <- markers %>% arrange(desc(stat)) %>% dplyr::select(gene, stat)

ranks <- deframe(ranks)
head(ranks)

```

#### GOBP_Fatty_acid_beta_oxidation

```{r}
plotEnrichment(Signatures[["GOBP_FATTY_ACID_BETA_OXIDATION"]], ranks) + labs(title = "GOBP_FATTY_ACID_BETA_OXIDATION")
ggsave("Fig_2M.png", path = path_to_save,width = 4, height = 2)
```

#### CSC signature

```{r}
plotEnrichment(Signatures[["CSC_Herreros_Pomares"]], ranks) + labs(title = "CSC Signature")
ggsave("Fig_S1F.png", path = path_to_save, width=4, height = 3)
```


### Bar Plot

#### Senescence signatures

```{r}
# Load pathways from MSigDB and select the ones related to senescence
m_df <- msigdbr(species = "Homo sapiens")
pathways <- m_df %>% split(x = .$gene_symbol, f = .$gs_name)
pathways_senescence<- list(
  "GOBP_REGULATION_OF_CELLULAR_SENESCENCE"= pathways$GOBP_REGULATION_OF_CELLULAR_SENESCENCE,
  "KAMMINGA_SENESCENCE"= pathways$KAMMINGA_SENESCENCE,
  "CHICAS_RB1_TARGETS_SENESCENT"= pathways$CHICAS_RB1_TARGETS_SENESCENT,
  "GOBP_CELLULAR_SENESCENCE"= pathways$GOBP_CELLULAR_SENESCENCE,
  "TANG_SENESCENCE_TP53_TARGETS_UP"= pathways$TANG_SENESCENCE_TP53_TARGETS_UP,
  "FRIDMAN_SENESCENCE_DN"= pathways$FRIDMAN_SENESCENCE_DN)

# Run the fgsea
fgseaRes <- fgsea(pathways_senescence, stats = ranks)

# Organize the table
fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))

# Plot the results 
ggplot(fgseaResTidy, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill= padj)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="SENESCENCE SIGNATURES") + 
  theme_minimal()
ggsave("Fig_S1G.png", path = path_to_save, height = 4)

rm(fgseaRes, fgseaResTidy, pathways, m_df, pathways_senescence)
```


#### GSEA Leading genes between Mtorc1 signalling, Glycolysis & Cholesterol homeostasis pathways

```{r}
# Load pathways from MSigDB and select the ones you want to visualize
m_df <- msigdbr(species = "Homo sapiens", category = "H")
pathways <- m_df %>% split(x = .$gene_symbol, f = .$gs_name)

hallmark_sets <- list(
  "HALLMARK_MTORC1_SIGNALING" = pathways$HALLMARK_MTORC1_SIGNALING,
  "HALLMARK_GLYCOLYSIS" = pathways$HALLMARK_GLYCOLYSIS,
  "HALLMARK_CHOLESTEROL_HOMEOSTASIS" = pathways$HALLMARK_CHOLESTEROL_HOMEOSTASIS)

# Convert to TERM2GENE format — ensure gene names are in second column
term2gene <- do.call(rbind, lapply(names(hallmark_sets), function(term) {
  data.frame(term = term, gene = hallmark_sets[[term]])
}))

# Run the GSEA
gsea_res <- GSEA(
  geneList = ranks,
  TERM2GENE = term2gene,  
  pvalueCutoff = 0.05,
  verbose = TRUE,
)

# Plot the results
heatplot(gsea_res, foldChange = ranks, showCategory = 5) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0)
ggsave("Fig_S1I.png", path = path_to_save, height=3, width=13)

rm(m_df, pathways, hallmark_sets, term2gene, gsea_res, ranks)
```

## Gene-level analysis
#### Persistence markers table 
```{r}
#Generate the marker list 
Idents(data)<-"treatment"
persistence_markers<- FindMarkers(data,ident.1 = "treated" ,ident.2 = "control" , only.pos = TRUE, logfc.threshold = 0.25, min.pct = 0.3)
persistence_markers<-  dplyr::filter(persistence_markers, p_val_adj<0.05)
persistence_markers$gene <- rownames(persistence_markers)

#Save the table of deregulated genes 
write_xlsx(persistence_markers, file.path(path_to_save, "Table_S4.xlsx"))

rm(persistence_markers)

```


#### Pemetrexed resistance-associated genes

```{r}
VlnPlot(data, features = c("DHFR", "GART","TYMS", "ABCC4", "SLC46A1" ), group.by = "treatment", stack = TRUE, flip = TRUE) 
ggsave("Fig_2H1.png", path = path_to_save,width = 4, height = 4)
```

#### Cisplatin resistance-associated genes

```{r}
VlnPlot(data, features = c( "GSS","ABCC2", "ERCC1", "ERCC4","CDKN1A"), group.by = "treatment", stack = TRUE, flip = TRUE) 
ggsave("Fig_2H2.png", path = path_to_save,width = 4, height = 4)
```

#### Persistence-associated genes

```{r}
VlnPlot(data, features = c("PINK1", "CD44","NOTCH3", "ALDH1A1", "HDAC1"), group.by = "treatment", stack = TRUE, flip = TRUE) 
ggsave("Fig_2J.png", path = path_to_save,width = 4, height = 4)
```

#### Apoptosis-related genes

```{r}
VlnPlot(data, features = c("BAX", "BID", "PMAIP1","DDIT3", "TXNIP", "RHOB"), group.by = "treatment", stack = TRUE, flip = TRUE) 
ggsave("Fig_2K.png", path = path_to_save,width = 4, height = 4)
```

#### Glycolysis and OXPHO-related genes

```{r}
VlnPlot(data, features = c("HK2", "ENO2", "NDUFC2", "UQCRQ", "ATP5MF"), group.by = "treatment", stack = TRUE, flip = TRUE)
ggsave("Fig_2L.png", path = path_to_save,width = 4, height = 5)
```

#### Fatty acid ß-oxidation-related genes

```{r}
VlnPlot(data, features = c("CPT1A", "DECR1"), group.by = "treatment", stack = TRUE, flip = TRUE)
ggsave("Fig_S1E.png", path = path_to_save,width = 5, height = 3)
```

#### Hallmark KRAS signaling DOWN

```{r}
# Create a list of the genes of the Hallmark KRAS signaling down signature present in the DEG upon treatment
markers_KRAS_SIGNALING_DN <-  markers %>% 
  filter(gene %in% Signatures[["HALLMARK_KRAS_SIGLING_DN"]])

# Select the 20 most upregulated genes upon treatment
top20_up_genes <- markers_KRAS_SIGNALING_DN%>%
  arrange(desc(avg_log2FC)) %>%  # Sort by decreasing log2FC
  slice_head(n = 20)
genes_to_plot <- top20_up_genes$gene

# Plot the expression of these genes in control and treated cells
DotPlot(data, 
        features = genes_to_plot, group.by = "treatment",  scale = T) +
  coord_flip() 
ggsave("Fig_S1H.png", path = path_to_save)

rm(markers_KRAS_SIGNALING_DN, top20_up_genes)
```

#### TK-RR Violin plot

```{r}
VlnPlot(data, features = c("TK1", "RRM1"), group.by = "treatment") 
ggsave("Fig_4B.png", path = path_to_save, height=4, width =6)
```


## Venn diagram of resistance and persistence genes and dataset's DEG
```{r}

# INPUT: Upload resistance- & persistence- associated genes from the supplementary data excel sheet 
df <- read_excel("/Users/fauveauc/Documents/R_studio/A549 analysis/Supplemental_tables.xlsx", sheet = "Table S.3")
resistance_genes <- df[[2]]
persistence_genes <- df[[1]]
resistance_genes <- resistance_genes[!is.na(resistance_genes)]
persistence_genes <- persistence_genes[!is.na(persistence_genes)]

# List of significantly deregulated genes (DEG)
Idents(data)<- "treatment"
markers2<- FindMarkers(data,ident.1 = "treated" ,ident.2 = "control" ,logfc.threshold = 0.025, min.pct = 0.3)
markers2<- markers2 %>% dplyr::filter(p_val_adj < 0.05) 

# Venn diagram of the overlap between resistance genes, persistence genes and DEG
png(file.path(path_to_save, "Fig_2I.png"))
venn.plot <- draw.triple.venn(
  area1 = length(persistence_genes),
  area2 = length(resistance_genes),
  area3 = length(rownames(markers2)),
  n12 = length(intersect(persistence_genes, resistance_genes)),
  n13 = length(intersect(persistence_genes,rownames(markers2))),
  n23 = length(intersect(resistance_genes, rownames(markers2))),
  n123 = length(Reduce(intersect, list(rownames(markers2), resistance_genes, persistence_genes))),
  category = c("Persistence", "Resistance", "DEG"), 
  fill = c("blue", "red", "green"),
   alpha = 0.5,
  cex = 1.5,
  cat.cex = 1.5,
  lwd = 2,
  cat.fontfamily = "Arial"
)

grid.newpage()
grid.draw(venn.plot)
dev.off()

rm(df, markers2, venn.plot, persistence_genes, resistance_genes)

```
