---
title: "Multidataset Comparison"
format:
  html:
    toc: true
    number-sections: true
execute:
  cache: false
  freeze: auto
editor: 
  markdown: 
    wrap: 72
---

# Package & object loading

```{r}
library(Matrix)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(cowplot)
library(SingleCellExperiment)
library(hdf5r)
library(hexbin)
library(tibble)
library(dplyr)
library(fgsea)
library(msigdbr)
library(reshape2)
library(ComplexHeatmap)
library (tidyr)
library(colorRamp2)
library(readxl)
library(clusterProfiler)
library(enrichplot)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(openxlsx2)
library(svglite)
library(ggplot2)
library(ggVennDiagram)
library(AUCell)


# INPUT : Load the A549 spheroid seurat object 
data <- readRDS(file = "/Users/fauveauc/Documents/R_studio/A549 analysis/seurat_object_A549_pers_final.rds")

# INPUT : Obtain signatures from the supplementary tables of the publication :
  #Load the excel file
df <- read_excel("/Users/fauveauc/Documents/R_studio/A549 analysis/Supplemental_tables.xlsx", sheet = "Table S.7")
  # transfrom it into a list containing the different signatures 
Signatures <- lapply(seq_len(nrow(df)), function(i) {
  g <- unlist(df[i, -1], use.names = FALSE)
  g <- as.character(g)
  g <- trimws(g)
  g <- g[!is.na(g) & g != ""]
  unique(g)
})
names(Signatures) <- df[[1]]

# OUTPUT: Determine the path to the directory where the figures and table will be saved
path_to_save <- "/Users/fauveauc/Documents/R_studio/A549 analysis/figures_and_tables"


rm(df)
```

# Preclinical datasets comparison

## Pathway comparison

### Datasets preparation for GSEA (no pvalue filtering)

```{r}
# Creation of a data frame containing the DEG of the A549 spheroids upon treatment
Idents(data)<-"treatment"
markers<- FindMarkers(data,ident.1 = "treated" ,ident.2 = "control" ,logfc.threshold = 0, min.pct = 0.3)
markers_dataset<- markers %>% dplyr::select(-p_val, -p_val_adj, -pct.1, -pct.2)%>% dplyr::rename(Rankvalue = avg_log2FC)
markers_dataset$Gene <- rownames(markers_dataset)

# Creation of DEG data frames from literature
##### Dhimolea et al. 

# INPUT : Load the excel Dhimolea et al. Supplementary table  
Dhimolea <- read_excel("/Volumes/projects/Persist-seq/Clara/bio-info/CAF crosstalk project/object_jed/01042025_comparison_with_lit/tables/Dhimolea_tableS1.xlsx")

# Extract datasets and format the data frames 
Dhimolea <- Dhimolea%>% distinct(symbol, .keep_all = TRUE) 
Dhimolea_HCI002<- Dhimolea %>% dplyr::select( HCI002_Org_Doc_vs_HCI002_Org_Veh_log2FoldChange, symbol)%>% dplyr::rename(Rankvalue = HCI002_Org_Doc_vs_HCI002_Org_Veh_log2FoldChange,Gene= symbol)

Dhimolea_PM154<- Dhimolea %>% dplyr::select( PM154_Org_Doc_vs_PM154_Org_Veh_log2FoldChange, symbol)%>% dplyr::rename(Rankvalue = PM154_Org_Doc_vs_PM154_Org_Veh_log2FoldChange,Gene= symbol)

###### Hangauer et al. 

# INPUT : Load the excel Hangauer et al. Supplementary table  
Hangauer <- read_excel("/Volumes/projects/Persist-seq/Clara/bio-info/CAF crosstalk project/object_jed/01042025_comparison_with_lit/tables/Hangauer_S1.xlsx")

# Format the data frame
Hangauer<- Hangauer %>% dplyr::select(-q_value, -`RefSeq IDs`, -`Parental FPKM`, -`Persister FPKM` )%>% dplyr::rename(Rankvalue = foldchange, Gene= `Gene ID`)
Hangauer <- Hangauer %>% distinct(Gene, .keep_all = TRUE) 
Hangauer$Rankvalue <- as.numeric(Hangauer$Rankvalue)
Hangauer$Rankvalue <- log2(Hangauer$Rankvalue)

###### Moghal et al. 

# INPUT : Load the excel Moghal et al. DEG list 
Moghal <- read_excel("/Volumes/projects/Persist-seq/Clara/bio-info/CAF crosstalk project/object_jed/01042025_comparison_with_lit/datasets/Moghal.xlsx")

# Format the data frame
Moghal<- Moghal %>% dplyr::select(Gene.symbol, logFC)%>% dplyr::rename(Rankvalue = `logFC`, Gene = `Gene.symbol`) 

######  Rehman et al.

# INPUT : Load the excel Rehman et al. Supplementary table  
Rehman <- read_excel("/Volumes/projects/Persist-seq/Clara/bio-info/CAF crosstalk project/object_jed/01042025_comparison_with_lit/tables/Rehman_S4.xlsx")

# Format the data frame
colnames(Rehman) <- Rehman[1, ]
Rehman<- Rehman %>% slice(-1)  %>% dplyr::select(`logFC(DTP/CTRL)`, GeneID) %>%dplyr::rename(Rankvalue = `logFC(DTP/CTRL)`, Gene=GeneID)
Rehman <- Rehman %>% distinct(Gene, .keep_all = TRUE) 


# Within DEG dataframes,  
#     keep only rows with genes name that match official gene symbols
#     remove duplicated genes
#     Ensure Rank values are numeric and different from NA or "-Inf" 
#     Rank values in descending order
valid_symbols <- keys(org.Hs.eg.db, keytype = "SYMBOL")
for (x in c("Rehman", "Moghal", "Dhimolea_HCI002", "Dhimolea_PM154",  "Hangauer", "markers_dataset")){
  y<- get(x)
  y <- y[y$Gene %in% valid_symbols,,drop = FALSE ]%>%
  distinct(Gene, .keep_all = TRUE) 
  y$Rankvalue <- as.numeric(y$Rankvalue)
  y <- y %>% filter(!is.na(as.numeric(Rankvalue))) %>%
  filter(Rankvalue != "-Inf")  
  y<- arrange(y, desc(Rankvalue))
  assign(x, y)
}

# Create a list containing the ranked genes of all the datasets 
list_datasets <- list(This_study=markers_dataset, Rehman=Rehman, Moghal=Moghal, Dhimolea_HCI002=Dhimolea_HCI002, Dhimolea_PM154=Dhimolea_PM154,Hangauer=Hangauer)

ranked_lists <- lapply(list_datasets, function(df) {
  setNames(df$Rankvalue, df$Gene)  })

rm(markers_dataset, Moghal, Rehman, y, x, valid_symbols, Hangauer, Dhimolea, Dhimolea_HCI002, Dhimolea_PM154)
```

### GSEA 

```{r}

# Load MSigDB Hallmark gene sets 
m_df <- msigdbr(species = "Homo sapiens", category = "H")
head(m_df, 2) %>% as.data.frame
fgsea_sets <- m_df %>% split(x = .$gene_symbol, f = .$gs_name)

# Add Ascierto signatures to the list of pathways 
fgsea_sets[["Ascierto_Up"]] <- Signatures[["Ascierto_UP"]]
fgsea_sets[["Ascierto_Down"]] <- Signatures[["Ascierto_DOWN"]]

# Run the gsea
gsea_results <- lapply(names(ranked_lists), function(dataset) {
  fgsea(pathways = fgsea_sets, stats = ranked_lists[[dataset]], nperm = 1000000000)})
names(gsea_results) <- names(ranked_lists)

# Convert results to a data frame
gsea_results_df <- do.call(rbind, Map(cbind, gsea_results, dataset = names(gsea_results)))

```
### Hallmarks heatmap
```{r}
# Heatmap preparation
#   Select only Hallmarks results 
gsea_results_df_hallmarks <- dplyr::filter(gsea_results_df,!pathway %in% c('Ascierto_Up', "Ascierto_Down"))

#   Create a data frame with pathways names and pvalues 
#   Remove pathway not significantly modified in at least one dataset and values with NA

heatmap_data_pvalue <- dcast(gsea_results_df_hallmarks, pathway ~ dataset, value.var = "padj", fill = 1)
heatmap_data_pvalue<- heatmap_data_pvalue[apply(heatmap_data_pvalue[, -1], 1, function(x) any(x < 0.05)), ]


#   Create a data frame with pathways names and NES 
#   Remove pathway not significantly modified in at least one dataset (using the heatmap_data_pvalue dataframe)
heatmap_data <- dcast(gsea_results_df_hallmarks, pathway ~ dataset, value.var = "NES", fill = 0)
heatmap_data <- heatmap_data[heatmap_data$pathway %in% heatmap_data_pvalue$pathway, ]


#   Convert to matrix
heatmap_matrix_NES <- as.matrix(heatmap_data[, -1])
rownames(heatmap_matrix_NES) <- heatmap_data$pathway
heatmap_matrix_pval <- as.matrix(heatmap_data_pvalue[, -1])
rownames(heatmap_matrix_pval) <- heatmap_data_pvalue$pathway

# Heatmap generation

svglite(file.path(path_to_save, "Fig_3B.svg"),width = 6, height = 7)
Heatmap(
  heatmap_matrix_NES,
  name = "NES",
  col = colorRamp2(c(min(heatmap_matrix_NES), 0, max(heatmap_matrix_NES)), c("blue", "white", "red")),
  cluster_rows = TRUE,
  cluster_columns = FALSE,  # Disable automatic column clustering
  column_order = c("This_study","Dhimolea_HCI002", "Dhimolea_PM154","Moghal","Rehman", "Hangauer"),
  show_column_names = TRUE,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 8),
  row_names_max_width = unit(20, "cm"),
  heatmap_legend_param = list(
    title = "NES",
    labels_gp = gpar(fontsize = 10, fontface = "italic"),
    legend_width = unit(15, "cm")
  ),
  cell_fun = function(j, i, x, y, w, h, fill) {
    if (!is.na(heatmap_matrix_pval[i, j])) {
      if (heatmap_matrix_pval[i, j] < 0.001) {
        grid.text("∗∗∗", x, y, gp = gpar(fontsize = 11, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.01) {
        grid.text("∗∗", x, y, gp = gpar(fontsize = 11, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.05) {
        grid.text("∗", x, y, gp = gpar(fontsize = 11, col = "black"))
      }
    }
  }
)
dev.off()

rm(fgsea_sets, gsea_results_df_hallmarks, heatmap_data, heatmap_data_pvalue, heatmap_matrix_NES, heatmap_matrix_pval, m_df, gsea_results, ranked_lists)

```

### Ascierto signatures heatmap

```{r}
# Heatmap preparation 
#   Selection of the Ascierto signature results
gsea_results_df_ascierto <- dplyr::filter(gsea_results_df,pathway %in% c('Ascierto_Up', "Ascierto_Down"))

#   Create a data frame with pathways names and pvalue 
heatmap_data_pvalue <- dcast(gsea_results_df_ascierto, pathway ~ dataset, value.var = "padj", fill = 1)

#   Create a data frame with pathways names and NES
heatmap_data <- dcast(gsea_results_df_ascierto, pathway ~ dataset, value.var = "NES", fill = 0)

# Convert to matrix
heatmap_matrix_NES <- as.matrix(heatmap_data[, -1])
rownames(heatmap_matrix_NES) <- heatmap_data$pathway
heatmap_matrix_pval <- as.matrix(heatmap_data_pvalue[, -1])
rownames(heatmap_matrix_pval) <- heatmap_data_pvalue$pathway

# Heatmap generation

svglite(file.path(path_to_save, "Fig_4A.svg"),width = 7, height = 4)
Heatmap(
  heatmap_matrix_NES,
  name = "NES",
  col = colorRamp2(c(min(heatmap_matrix_NES), 0, max(heatmap_matrix_NES)), c("blue", "white", "red")),
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_column_names = TRUE,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 8),  # Adjust row name size
  row_names_max_width = unit(20, "cm"),
  heatmap_legend_param = list(
    title = "NES",
    labels_gp = gpar(fontsize = 10, fontface = "italic"),  # Adjust legend label font
    legend_width = unit(15, "cm")   # Adjust legend width
  ),
    cell_fun = function(j, i, x, y, w, h, fill) {
   # Check for NA values first
    
      # Add asterisks based on p-value thresholds
      if (heatmap_matrix_pval[i, j] < 0.001) {
        grid.text("∗∗∗", x, y, gp = gpar(fontsize = 30, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.01) {
        grid.text("∗∗", x, y, gp = gpar(fontsize = 30, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.05) {
        grid.text("∗", x, y, gp = gpar(fontsize = 30, col = "black"))
      
    }
  }
 
)

dev.off()

rm(gsea_results_df_ascierto,gsea_results_df, heatmap_data, heatmap_data_pvalue, heatmap_matrix_NES, heatmap_matrix_pval)
```

## DTP signature identification

### Datasets preparation (filtering out non-significative DEG)
Same as previously except this time the DEGs are filtered (pvalue < 0.05) and abs(Rankvalue)>0.1 

```{r}
# Creation of a data frame containing the significantly DEG of the A549 spheroids upon treatment
markers_dataset<- markers %>% dplyr::filter(p_val_adj<0.05)
# Format the data frame
markers_dataset <-markers_dataset%>% dplyr::select(-p_val, -p_val_adj, -pct.1, -pct.2)%>% dplyr::rename(Rankvalue = avg_log2FC)
markers_dataset$Gene <- rownames(markers_dataset)

# Creation of DEG data frames from literature
# Dhimolea et al. 

# INPUT : Load the excel Dhimolea et al. Supplementary table  
Dhimolea <- read_excel("/Volumes/projects/Persist-seq/Clara/bio-info/CAF crosstalk project/object_jed/01042025_comparison_with_lit/tables/Dhimolea_tableS1.xlsx")

# Extract datasets and format the data frames 
Dhimolea <- Dhimolea%>% distinct(symbol, .keep_all = TRUE) 
Dhimolea_HCI002<- Dhimolea %>% dplyr::select( HCI002_Org_Doc_vs_HCI002_Org_Veh_log2FoldChange, symbol)%>% dplyr::rename(Rankvalue = HCI002_Org_Doc_vs_HCI002_Org_Veh_log2FoldChange,Gene= symbol)

Dhimolea_PM154<- Dhimolea %>% dplyr::select( PM154_Org_Doc_vs_PM154_Org_Veh_log2FoldChange, symbol)%>% dplyr::rename(Rankvalue = PM154_Org_Doc_vs_PM154_Org_Veh_log2FoldChange,Gene= symbol)

# Hangauer et al. 

# INPUT : Load the excel Hangauer et al. Supplementary table 
Hangauer <- read_excel("/Volumes/projects/Persist-seq/Clara/bio-info/CAF crosstalk project/object_jed/01042025_comparison_with_lit/tables/Hangauer_S1.xlsx")

# Format the data frame
Hangauer<- Hangauer  %>% dplyr::filter(q_value<0.05 )%>% dplyr::select(-q_value, -`RefSeq IDs`, -`Parental FPKM`, -`Persister FPKM` )%>% dplyr::rename(Rankvalue = foldchange, Gene= `Gene ID`)
Hangauer <- Hangauer %>% distinct(Gene, .keep_all = TRUE) 
Hangauer$Rankvalue <- as.numeric(Hangauer$Rankvalue)
Hangauer$Rankvalue <- log2(Hangauer$Rankvalue)

# Moghal et al. 

# INPUT : Load the excel Moghal DEG list
Moghal <- read_excel("/Volumes/projects/Persist-seq/Clara/bio-info/CAF crosstalk project/object_jed/01042025_comparison_with_lit/datasets/Moghal.xlsx")

# Format the data frame
Moghal$adj.P.Val <- as.numeric(Moghal$adj.P.Val)
Moghal<- Moghal %>% dplyr::filter(adj.P.Val<0.05) %>% dplyr::select(Gene.symbol, logFC)%>% dplyr::rename(Rankvalue = `logFC`, Gene = `Gene.symbol`) 

# Rehman et al. 

# INPUT : Load the excel Rehman et al. Supplementary table 
Rehman <- read_excel("/Volumes/projects/Persist-seq/Clara/bio-info/CAF crosstalk project/object_jed/01042025_comparison_with_lit/tables/Rehman_S4.xlsx")

# Format the data frame
colnames(Rehman) <- Rehman[1, ]
Rehman<- Rehman %>% slice(-1)  %>% dplyr::select(`logFC(DTP/CTRL)`, GeneID) %>%dplyr::rename(Rankvalue = `logFC(DTP/CTRL)`, Gene=GeneID)
Rehman <- Rehman %>% distinct(Gene, .keep_all = TRUE) 

# Upload the list of official gene symbols 
list_gene_symbols <- read_excel("/Users/fauveauc/Documents/R_studio/A549 analysis/list_symbols_names.xlsx")

# Within DEG dataframes,  
#     keep only rows with genes name that match official gene symbols
#     remove duplicated genes
#     Ensure Rank values are numeric and different from NA or "-Inf" 
#     Rank values in descending order
#.    keep only genes with abs(Rankvalue)>0.1 
for (x in c("Rehman", "Moghal", "Dhimolea_HCI002", "Dhimolea_PM154", "Hangauer", "markers_dataset")){
  y<- get(x)
  y <- y[y$Gene %in% list_gene_symbols$valid_symbols,,drop = FALSE ]
  y <- y[!duplicated(y$Gene), ]
  y$Rankvalue <- as.numeric(y$Rankvalue)
  y <- y %>% filter(!is.na(as.numeric(Rankvalue))) %>%
  filter(Rankvalue != "-Inf") 
  y<- arrange(y, desc(Rankvalue)) %>% filter(abs(Rankvalue)>0.1)
  assign(x, y)
}

list_datasets <- list(This_study=markers_dataset, Rehman=Rehman, Moghal=Moghal,  Dhimolea_PM154=Dhimolea_PM154, Dhimolea_HCI002=Dhimolea_HCI002, Hangauer=Hangauer )


rm(markers_dataset, Moghal, Rehman, y, x, valid_symbols, Hangauer, Dhimolea, Dhimolea_HCI002, Dhimolea_PM154)
```

### Identification of the conserved DEGs

```{r}

# Separation of genes up and down-regulated genes across datasets 
up_genes_list <- lapply(list_datasets, function(dataset) {
  as.character(dataset$Gene[dataset$Rankvalue > 0 ]) })

down_genes_list <- lapply(list_datasets, function(dataset) {
 as.character(dataset$Gene[dataset$Rankvalue < 0 ]) })

# Plot the Venn diagram of Upregulated genes
ggVennDiagram(up_genes_list, set_size = 5, label = "count", label_size = 3) + ggtitle("Upregulated Genes")+scale_fill_gradient(low = "#FF6B6B", high="#490000")+ scale_x_continuous(expand = expansion(mult = .2))
ggsave("Fig_3C1A.png", path = path_to_save, width=6, height = 6)
ggVennDiagram(up_genes_list, set_size = 0, label = "count", label_size = 0) + ggtitle("Upregulated Genes")+scale_fill_gradient(low = "#FF6B6B", high="#490000")+ scale_x_continuous(expand = expansion(mult = .2))
ggsave("Fig_3C1B.png", path = path_to_save, width=6, height = 6)

# Plot the Venn diagram of downregulated genes
ggVennDiagram(down_genes_list, set_size = 5, label = "count", label_size = 3) + ggtitle("Downregulated Genes") + scale_x_continuous(expand = expansion(mult = .2))
ggsave("Fig_3C2A.png", path = path_to_save, width=6, height = 6)
ggVennDiagram(down_genes_list, set_size = 0, label = "count", label_size = 0) + ggtitle("Downregulated Genes")+ scale_x_continuous(expand = expansion(mult = .2))
ggsave("Fig_3C2B.png", path = path_to_save, width=6, height = 6)


# Create a vector containing the conserved genes
common_up_genes <- Reduce(intersect, up_genes_list)
common_down_genes <- Reduce(intersect, down_genes_list)

# Save the list of conserved genes 
n <- max(length(common_up_genes), length(common_down_genes))
df <- data.frame(
  DTP_UP_SIGNATURE = c(common_up_genes, rep(NA_character_, n - length(common_up_genes))),
  DTP_DOWN_SIGNATURE  = c(common_down_genes,  rep(NA_character_, n - length(common_down_genes))))
write_xlsx(df, file.path(path_to_save, "Table_S6.xlsx"))

rm(get_top_up_down,top_lists,up_genes_list, down_genes_list, df,n)
```

### Visualization of the conserved gene expression in the different datasets

```{r}
# Helper function to extract fold changes
extract_fc <- function(genes, dataset_list) {
  fc_table <- data.frame(Gene = genes)
  
  for (name in names(dataset_list)) {
    dataset <- dataset_list[[name]]
    dataset <- dataset[dataset$Gene %in% genes, ]
    fc_table[[name]] <- dataset$Rankvalue[match(fc_table$Gene, dataset$Gene)]
  }
  
  return(fc_table)
}

# Create fold-change tables
up_fc_df <- extract_fc(common_up_genes, list_datasets)
down_fc_df <- extract_fc(common_down_genes, list_datasets)

############ Heatmap for genes of the DTP signature Up 
# Order the genes of the data.table and then  convert it into a matrix
up_fc_df[ , -1] <- lapply(up_fc_df[ , -1], function(x) as.numeric(as.character(x))) 
up_fc_df$mean_fc <- rowMeans(up_fc_df[ , -1], na.rm = TRUE)
up_fc_df <- up_fc_df[order(up_fc_df$mean_fc, decreasing = F), ]
up_fc_df$mean_fc <- NULL
rownames(up_fc_df) <- NULL
fc_matrix <- up_fc_df %>% column_to_rownames(var = 'Gene')%>%as.matrix()

# Plot as heatmap
col_fun <- colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))
png(file.path(path_to_save, "Fig_3D1.png"), width = 1400, height = 300, res = 150)
pheatmap(t(fc_matrix),
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         color = col_fun,
         fontsize_row = 10,
         fontsize_col = 8,
         scale="none")
dev.off()

############ Heatmap for genes of the DTP signature Down 
# Order the genes of the data.table and then  convert it into a matrix
down_fc_df[ , -1] <- lapply(down_fc_df[ , -1], function(x) as.numeric(as.character(x))) 
down_fc_df$mean_fc <- rowMeans(down_fc_df[ , -1], na.rm = TRUE)
down_fc_df <- down_fc_df[order(down_fc_df$mean_fc, decreasing = F), ]
down_fc_df$mean_fc <- NULL
rownames(down_fc_df) <- NULL
fc_matrix <- down_fc_df %>% column_to_rownames(var = 'Gene')%>%as.matrix()


# Plot as heatmap
png(file.path(path_to_save, "Fig_3D2.png"), width = 2000, height = 300, res = 150)
pheatmap(t(fc_matrix),
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         color = col_fun,
         fontsize_row = 10,
         fontsize_col = 7,
         scale="none")
dev.off()


rm(up_fc_df,down_fc_df, fc_matrix, list_datasets, col_fun, extract_fc, data)
```

# Comparison to patient NSCLC scRNA-seq 

## Preparation of the scRNA-seq dataset
The scRNA-seq dataset from Maynard et al., 2020 was obtained from the github repository czbiohub/scell_lung_adenocarcinoma (Data_input/objects/NI02_Epi_Seurat_object_clustered.Rdata)

```{r}
# INPUT : Enter the path to the scRNA-seq Rdata file containing the Seurat object of the NSCLC patient biopsies 
rdata_path <- "/Users/fauveauc/Documents/R_studio/A549 analysis/NI02_Epi_Seurat_object_clustered.Rdata"

# Load the Rdata into a clean env 
e <- new.env(parent = emptyenv())
load(rdata_path, envir = e)
ls(envir = e)

# Identify the seurat object
is_seurat <- vapply(ls(envir = e), function(nm) inherits(e[[nm]], "Seurat"), logical(1))
seurat_names <- names(is_seurat)[is_seurat]

# Load the seurat object 
data_patient <- e[[seurat_names]]
data_patient <- UpdateSeuratObject(data_patient)

#Remove non-cancerous samples ("NAT")
cells_to_keep <- rownames(data_patient@meta.data)[data_patient$sample_type != "NAT" | is.na(data_patient$sample_type)]
data_patient <- subset(data_patient, cells = cells_to_keep)
data_patient <- subset(data_patient, !is.na(biopsy_time_status))

#Add a metadata column named "recist" indicating the RECIST score at the time of biopsy 
data_patient$recist <- data_patient$biopsy_time_status
data_patient$recist <-ifelse(data_patient$biopsy_time_status == "Early", data_patient$early_treatment_status, data_patient$recist)
data_patient$recist <-ifelse(data_patient$biopsy_time_status == "off treatment", "naive", data_patient$recist)
data_patient$recist <- ifelse(data_patient$recist %in% c("PR","SD"), "PR/SD", data_patient$recist )
DimPlot(data_patient, reduction = "tsne", group.by = "recist")
data_patient$recist <- factor(
  data_patient$recist,
  levels = c("naive", "PR/SD", "PD")
)
DimPlot(data_patient,reduction = "tsne",group.by = "recist",cols = c("#6F9BF8", "#E87D72", "#53B74C" ))
ggsave( "Fig_3E.png", path = path_to_save, width=4, height = 3)

#Add a metadata column named "recist_timing" indicating the RECIST score and timepoint (early or late) at the time of biopsy 
data_patient$recist_timing <- paste(data_patient$recist, data_patient$biopsy_timing, sep = "_")

rm(rdata_path, e, is_seurat, cells_to_keep, seurat_names)
```

## UMAP with the AUCell score of the DTP signatures

```{r}
expr_mat <- GetAssayData(data_patient, assay = "RNA")

# Build AUCell rankings
expr_ranks <- AUCell_buildRankings(expr_mat)

# Calculate AUCell score for the pathway gene set

dtp_up_auc <- AUCell_calcAUC(list(DTP_signature_UP =  common_up_genes), expr_ranks)
dtp_down_auc <- AUCell_calcAUC(list(DTP_signature_DOWN = common_down_genes), expr_ranks)

#  Extract the score for each cell
dtp_up_scores <- as.numeric(getAUC(dtp_up_auc)["DTP_signature_UP", ])
dtp_down_scores <- as.numeric(getAUC(dtp_down_auc)["DTP_signature_DOWN", ])

# Add the score to Seurat metadata
data_patient$DTP_signature_UP <- dtp_up_scores
data_patient$DTP_signature_DOWN <- dtp_down_scores


# Plot on UMAP
FeaturePlot(
  data_patient,
  features = "DTP_signature_DOWN",
  reduction = "tsne"
)+ scale_color_gradientn(colors = c("grey90", "#B3D9FF", "#0073E6", "#001F4D"))
ggsave( "Fig_3F1.png", path = path_to_save, width=4, height = 3)
FeaturePlot(
  data_patient,
  features = "DTP_signature_UP",
  reduction = "tsne"
)+ scale_color_gradientn(colors = c("lightgrey", "pink", "red", "darkred"))
ggsave("Fig_3F2.png", path = path_to_save, width=4, height = 3)

rm(expr_mat, expr_ranks, dtp_up_auc, dtp_down_auc, dtp_up_scores, dtp_down_scores)
```

## Heatmap of DTP signatures enrichment

### Comparison of PD, PR/SD and all recist scores combined (figure 3.F)

```{r}
# Generation of the DEG for the GSEA
Idents(data_patient)<-"recist"
markers_tr<- FindMarkers(data_patient,ident.1 = c("PR/SD", "PD") ,ident.2 = "naive" ,logfc.threshold = 0, min.pct = 0.3)
markers_PD<- FindMarkers(data_patient,ident.1 =  "PD" ,ident.2 = "naive" ,logfc.threshold = 0, min.pct = 0.3)
markers_PRSD<- FindMarkers(data_patient,ident.1 = "PR/SD" ,ident.2 = "naive" ,logfc.threshold = 0, min.pct = 0.3)

list_comparisons <-c( "markers_PD","markers_PRSD","markers_tr")


# Preparation of the ranked list of genes for the GSEA
ranked_list <- lapply(list_comparisons, function(dataset_name) {
  # Creates a temporary dataframe containing for one dataset, one column with genes and  one with associated forlchanges
  temp_df <- get(dataset_name)
  temp_df$gene <- rownames(temp_df)
  temp_df$foldchange <- temp_df$avg_log2FC
  
  # Convert foldchange to numeric
  temp_df$foldchange <- as.numeric(as.character(temp_df$foldchange))
  
  # Remove rows with NA 
  temp_df <- temp_df[!is.na(temp_df$foldchange), ]
  
  #Remove duplicate Genes
  temp_df <- temp_df %>%  distinct(gene, .keep_all = TRUE) 
  
  # Sort by foldchange in descending order
  temp_df <- temp_df[order(-temp_df$foldchange), ]
  
  # Return a named list with genes and values
  list(
    genes = temp_df$gene,
    foldchange = temp_df$foldchange
  )
})
# Add dataset names as list names
names(ranked_list) <- list_comparisons

# Prepare the pathways
# Load gene sets 
fgsea_sets <- list()
m_df <- msigdbr(species = "Homo sapiens", category = "H")
head(m_df, 2) %>% as.data.frame
fgsea_sets <- m_df %>% split(x = .$gene_symbol, f = .$gs_name)
fgsea_sets[["DTP_Signature_Down"]] <- common_down_genes
fgsea_sets[["DTP_Signature_Up"]] <- common_up_genes

# GSEA
gsea_results <- lapply(names(ranked_list), function(dataset) {
   stats_vector <- setNames(ranked_list[[dataset]]$foldchange, ranked_list[[dataset]]$genes)
  fgsea(pathways = fgsea_sets, stats = stats_vector)
})
names(gsea_results) <- names(ranked_list)

# Convert results to a data frame
gsea_results_df <- do.call(rbind, Map(cbind, gsea_results, dataset = names(gsea_results)))

#  Heatmap 
#Selection of the results for the DTP signature pathways
gsea_results_df_DTP <- dplyr::filter(gsea_results_df,pathway %in% c('DTP_Signature_Down', "DTP_Signature_Up"))

# Data table preparation
# Create a data frame with patwhays and NES 
heatmap_data_pvalue <- dcast(gsea_results_df_DTP, dataset ~ pathway,, value.var = "padj", fill = 1)

heatmap_data <- dcast(gsea_results_df_DTP, dataset ~ pathway,, value.var = "NES", fill = 0)

# Convert to matrix
heatmap_matrix_NES <- as.matrix(heatmap_data[, -1])
rownames(heatmap_matrix_NES) <- heatmap_data$dataset
heatmap_matrix_pval <- as.matrix(heatmap_data_pvalue[, -1])
rownames(heatmap_matrix_pval) <- heatmap_data_pvalue$dataset

svglite(file.path(path_to_save, "Fig_3G.svg"),width = 5, height = 4)
Heatmap(
  heatmap_matrix_NES,
  name = "NES",
  col = colorRamp2(c(min(heatmap_matrix_NES), 0, max(heatmap_matrix_NES)), c("blue", "white", "red")),
  cluster_rows = T,
  cluster_columns = F,
  show_column_names = TRUE,
  show_row_names = TRUE,
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 8),  # Adjust row name size
  row_names_max_width = unit(20, "cm"),
  heatmap_legend_param = list(
    title = "NES",
    labels_gp = gpar(fontsize = 10, fontface = "italic"),  # Adjust legend label font
    legend_width = unit(15, "cm")   # Adjust legend width
  ),
    cell_fun = function(j, i, x, y, w, h, fill) {
   # Check for NA values first
    
      # Add asterisks based on p-value thresholds
      if (heatmap_matrix_pval[i, j] < 0.001) {
        grid.text("∗∗∗", x, y, gp = gpar(fontsize = 35, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.01) {
        grid.text("∗∗", x, y, gp = gpar(fontsize = 35, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.05) {
        grid.text("∗", x, y, gp = gpar(fontsize = 35, col = "black"))
      
    }
  }
 
)
dev.off()
rm(markers_tr, markers_PD, markers_PRSD, list_comparisons, ranked_list, gsea_results_df, gsea_results, gsea_results_df_DTP, heatmap_data, heatmap_data_pvalue, heatmap_matrix_NES, heatmap_matrix_pval)
```

### Comparison PD, PR/SD across timepoints (figure 3.H)

```{r}
# Generation of the DEG for the GSEA
Idents(data_patient)<-"recist_timing"
markers_early_naivevsPD<- FindMarkers(data_patient,ident.1 = "PD_Early" ,ident.2 = "naive_Baseline" ,logfc.threshold = 0, min.pct = 0.3)
markers_early_naivevsPRSD<- FindMarkers(data_patient,ident.1 = "PR/SD_Early" ,ident.2 = "naive_Baseline" ,logfc.threshold = 0, min.pct = 0.3)

markers_late_naivevsPD<- FindMarkers(data_patient,ident.1 = "PD_Late" ,ident.2 = "naive_Baseline" ,logfc.threshold = 0, min.pct = 0.3)
markers_late_naivevsPRSD<- FindMarkers(data_patient,ident.1 = "PR/SD_Late" ,ident.2 = "naive_Baseline" ,logfc.threshold = 0, min.pct = 0.3)

list_comparisons <-c("markers_early_naivevsPD", "markers_early_naivevsPRSD", "markers_late_naivevsPD", "markers_late_naivevsPRSD")


# Preparation of the ranked list of genes for the GSEA
ranked_list <- lapply(list_comparisons, function(dataset_name) {
  # Creates a temporary dataframe containing for one dataset, one column with genes and  one with associated forlchanges
  temp_df <- get(dataset_name)
  temp_df$gene <- rownames(temp_df)
  temp_df$foldchange <- temp_df$avg_log2FC
  
  # Convert foldchange to numeric
  temp_df$foldchange <- as.numeric(as.character(temp_df$foldchange))
  
  # Remove rows with NA 
  temp_df <- temp_df[!is.na(temp_df$foldchange), ]
  
  # Remove duplicate Genes
  temp_df <- temp_df %>%  distinct(gene, .keep_all = TRUE) 
  
  # Sort by foldchange in descending order
  temp_df <- temp_df[order(-temp_df$foldchange), ]
  
  # Return a named list with genes and values
  list(
    genes = temp_df$gene,
    foldchange = temp_df$foldchange
  )
})
# Add dataset names as list names
names(ranked_list) <- list_comparisons


# GSEA
gsea_results <- lapply(names(ranked_list), function(dataset) {
   stats_vector <- setNames(ranked_list[[dataset]]$foldchange, ranked_list[[dataset]]$genes)
  fgsea(pathways = fgsea_sets, stats = stats_vector)
})
names(gsea_results) <- names(ranked_list)

# Convert results to a data frame
gsea_results_df <- do.call(rbind, Map(cbind, gsea_results, dataset = names(gsea_results)))

# Heatmap 
# Selection of the results for the DTP signature pathways
gsea_results_df_DTP <- dplyr::filter(gsea_results_df,pathway %in% c('DTP_Signature_Down', "DTP_Signature_Up"))

# Data table preparation
# Create a data frame with patwhays and NES 
heatmap_data_pvalue <- dcast(gsea_results_df_DTP, dataset ~ pathway,, value.var = "padj", fill = 1)

heatmap_data <- dcast(gsea_results_df_DTP, dataset ~ pathway,, value.var = "NES", fill = 0)

# Convert to matrix
heatmap_matrix_NES <- as.matrix(heatmap_data[, -1])
rownames(heatmap_matrix_NES) <- heatmap_data$dataset
heatmap_matrix_pval <- as.matrix(heatmap_data_pvalue[, -1])
rownames(heatmap_matrix_pval) <- heatmap_data_pvalue$dataset

svglite(file.path(path_to_save, "Fig_3H.svg"),width = 5, height = 4)
Heatmap(
  heatmap_matrix_NES,
  name = "NES",
  col = colorRamp2(c(min(heatmap_matrix_NES), 0, max(heatmap_matrix_NES)), c("blue", "white", "red")),
  cluster_rows = T,
  cluster_columns = F,
  show_column_names = TRUE,
  show_row_names = TRUE,
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 8),  # Adjust row name size
  row_names_max_width = unit(20, "cm"),
  heatmap_legend_param = list(
    title = "NES",
    labels_gp = gpar(fontsize = 10, fontface = "italic"),  # Adjust legend label font
    legend_width = unit(15, "cm")   # Adjust legend width
  ),
    cell_fun = function(j, i, x, y, w, h, fill) {
   # Check for NA values first
    
      # Add asterisks based on p-value thresholds
      if (heatmap_matrix_pval[i, j] < 0.001) {
        grid.text("∗∗∗", x, y, gp = gpar(fontsize = 35, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.01) {
        grid.text("∗∗", x, y, gp = gpar(fontsize = 35, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.05) {
        grid.text("∗", x, y, gp = gpar(fontsize = 35, col = "black"))
      
    }
  }
 
)
dev.off()

rm(markers_early_naivevsPD, markers_early_naivevsPRSD, markers_late_naivevsPD, markers_late_naivevsPRSD, list_comparisons, ranked_list, gsea_results_df, gsea_results, gsea_results_df_DTP, heatmap_data, heatmap_data_pvalue, heatmap_matrix_NES, heatmap_matrix_pval, data_patient)
```

# Comparison to bulk MRD datasets

## Data preparation

```{r}
# INPUT : Enter the path to the Supplementary table 2 of Liu et al., 2024.
Liu_MRD_dataframe <- read_excel("/Volumes/projects/Persist-seq/Clara/bio-info/CAF crosstalk project/object_jed/01042025_comparison_with_lit/table_patient_sig/Liu_40164_2024_509_MOESM2_ESM.xlsx")
Liu_MRD_dataframe <- Liu_MRD_dataframe[-1, ]

# Creation of a list containing ranked DEG of each datasets
ranked_list <- lapply(names(Liu_MRD_dataframe)[-1], function(dataset_name) {
  # Creates a temporary dataframe containing for one dataset, one column with genes and  one with associated forlchanges
  temp_df <- Liu_MRD_dataframe[, c("gene", dataset_name)]
  colnames(temp_df) <- c("gene", "foldchange")
  
  # Convert foldchange to numeric
  temp_df$foldchange <- as.numeric(as.character(temp_df$foldchange))
  
  # Remove rows with NA 
  temp_df <- temp_df[!is.na(temp_df$foldchange), ]
  
  #Remove duplicate Genes
  temp_df <- temp_df %>%  distinct(gene, .keep_all = TRUE) 
  
  # Sort by foldchange in descending order
  temp_df <- temp_df[order(-temp_df$foldchange), ]
  
  # Return a named list with genes and values
  list(
    genes = temp_df$gene,
    foldchange = temp_df$foldchange
  )
})

# Add dataset names as list names
names(ranked_list) <- names(Liu_MRD_dataframe)[-1]

#Uniformization of the dataset names 
names(ranked_list) <- sub("_BR$", "_BRCA", names(ranked_list))
names(ranked_list) <- sub("_BRC$", "_BRCA", names(ranked_list))
names(ranked_list) <- sub("-BRC$", "_BRCA", names(ranked_list))
names(ranked_list) <- sub("_Re$", "_READ", names(ranked_list))
names(ranked_list) <- sub("_Rec$", "_READ", names(ranked_list))
names(ranked_list) <- sub("_CO$", "_COAD", names(ranked_list))


rm(Liu_MRD_dataframe)
```

## GSEA

```{r}
# Add Ascierto's signature to the list of pathways
fgsea_sets[["Ascierto_Up"]] <- Signatures[["Ascierto_UP"]]
fgsea_sets[["Ascierto_Down"]] <- Signatures[["Ascierto_DOWN"]]

# Run GSEA for each dataset
gsea_results <- lapply(names(ranked_list), function(dataset) {
   stats_vector <- setNames(ranked_list[[dataset]]$foldchange, ranked_list[[dataset]]$genes)
  fgsea(pathways = fgsea_sets, stats = stats_vector)
})
names(gsea_results) <- names(ranked_list)

# Convert results to a data frame
gsea_results_df <- do.call(rbind, Map(cbind, gsea_results, dataset = names(gsea_results)))

rm(ranked_list, m_df, fgsea_sets, gsea_results, common_up_genes, common_down_genes)
```

## Heatmap of DTP signatures enrichment

```{r}
#Selection of the results for the DTP signature pathways
gsea_results_df_DTP <- dplyr::filter(gsea_results_df,pathway %in% c('DTP_Signature_Down', "DTP_Signature_Up"))

# Data table preparation
#  Create a data frames with pathways and either NES score or p_value
heatmap_data_pvalue <- dcast(gsea_results_df_DTP, dataset ~ pathway,, value.var = "padj", fill = 1)

heatmap_data <- dcast(gsea_results_df_DTP, dataset ~ pathway,, value.var = "NES", fill = 0)

# Convert to matrix
heatmap_matrix_NES <- as.matrix(heatmap_data[, -1])
rownames(heatmap_matrix_NES) <- heatmap_data$dataset
heatmap_matrix_pval <- as.matrix(heatmap_data_pvalue[, -1])
rownames(heatmap_matrix_pval) <- heatmap_data_pvalue$dataset

# Heatmap generation
svglite(file.path(path_to_save, "Fig_3I.svg"))
Heatmap(
  heatmap_matrix_NES,
  name = "NES",
  col = colorRamp2(c(-3, 0, 2), c("blue", "white", "red")),
  cluster_rows = T,
  cluster_columns = F,
  show_column_names = TRUE,
  show_row_names = TRUE,
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 8),  # Adjust row name size
  row_names_max_width = unit(20, "cm"),
  heatmap_legend_param = list(
    title = "NES",
    labels_gp = gpar(fontsize = 10, fontface = "italic"),  # Adjust legend label font
    legend_width = unit(15, "cm")   # Adjust legend width
  ),
    cell_fun = function(j, i, x, y, w, h, fill) {
   # Check for NA values first
    
      # Add asterisks based on p-value thresholds
      if (heatmap_matrix_pval[i, j] < 0.001) {
        grid.text("∗∗∗", x, y, gp = gpar(fontsize = 30, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.01) {
        grid.text("∗∗", x, y, gp = gpar(fontsize = 30, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.05) {
        grid.text("∗", x, y, gp = gpar(fontsize = 30, col = "black"))
      
    }
  }
 
)
dev.off()

```

## Ascierto signature Heatmap

```{r}
# Data preparation 
#  Selection of the results with the Ascierto signature
gsea_results_df_ascierto <- dplyr::filter(gsea_results_df,pathway %in% c('Ascierto_Up', "Ascierto_Down"))
gsea_results_df_ascierto$dataset <- as.factor(gsea_results_df_ascierto$dataset)


#Data table preparation
#  Create a data frames with pathways and either NES score or p_value
heatmap_data_pvalue <- dcast(gsea_results_df_ascierto, pathway ~ dataset, value.var = "padj", fill = 1)

heatmap_data <- dcast(gsea_results_df_ascierto, pathway ~ dataset, value.var = "NES", fill = 0)

# Convertion to matrix
heatmap_matrix_NES <- as.matrix(heatmap_data[, -1])
rownames(heatmap_matrix_NES) <- heatmap_data$pathway
heatmap_matrix_pval <- as.matrix(heatmap_data_pvalue[, -1])
rownames(heatmap_matrix_pval) <- heatmap_data_pvalue$pathway

# Heatmap generation
svglite(file.path(path_to_save, "Fig_S4A.svg"),width = 7, height = 4)
Heatmap(
  heatmap_matrix_NES,
  name = "NES",
  col = colorRamp2(c(min(heatmap_matrix_NES), 0, max(heatmap_matrix_NES)), c("blue", "white", "red")),
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_column_names = TRUE,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 8),  # Adjust row name size
  row_names_max_width = unit(20, "cm"),
  heatmap_legend_param = list(
    title = "NES",
    labels_gp = gpar(fontsize = 10, fontface = "italic"),  # Adjust legend label font
    legend_width = unit(15, "cm")   # Adjust legend width
  ),
    cell_fun = function(j, i, x, y, w, h, fill) {
   # Check for NA values first
    
      # Add asterisks based on p-value thresholds
      if (heatmap_matrix_pval[i, j] < 0.001) {
        grid.text("∗∗∗", x, y, gp = gpar(fontsize = 12, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.01) {
        grid.text("∗∗", x, y, gp = gpar(fontsize = 12, col = "black"))
      } else if (heatmap_matrix_pval[i, j] < 0.05) {
        grid.text("∗", x, y, gp = gpar(fontsize = 12, col = "black"))
      
    }
  }
 
)
dev.off()

rm(gsea_results_df_ascierto, gsea_results_df, heatmap_data, heatmap_data_pvalue, heatmap_matrix_NES, heatmap_matrix_pval)
```
